working title Compatibility system and stygma size of pollen recipient as main predictors of heterospecific pollen effect

### 2018-12-04 Manuscript_draft/Manuscript_draft.Rmd
last edits of today

Recent studies have advanced in the ecological understanding of heterospecific pollen effect @morales2008 @ashman2013 @arceo2016. A general overview of foreign pollen arrival is that it can play an important role on species fitness but seems to be context dependent and not always produce a decrease in fitness @morales2008. Part of this unpredictability is due to the enormous variability of foreing pollen transferred in nature, where levels between 0 and 75 percent are seen, but most commonly values ranges between 10 and 20 percent of the total pollen load, being the generalist species the ones that receive greater loads of heterospecific pollen @montgomery2012 @fang2013. Although heterospecific pollen quantity is fundamental to understand the outcome of the interaction so is the different traits of both pollen donor and recipient. @ashman2013 postulated the first predictive framework for traits of heterospecific pollen effect, where different traits such as compatibility system and pollen size among others seems to be crucial to understand foreing pollen effect. Moreover, in @tong2016 a assymetric effect was shown in a crossing experiment between 6 species of the genus *Pedicularis* where the pollen of long styled species was able to grow the full length of the style on short styled species but not viceversa. Despite these recent caveats, we still lack empirical evidence to affirm what are the main traits that drive heterospecific pollen effect for both pollen donor and recipient at seed production level. Interestingly, to comprehend how this traits interact is also crucial to look at the phylogenetic relatedness of the species. There is a considerable amount of literature of crosses between close related species @brown2001 @arceo2016can @tong2016 for which is predicted to have a greater negative impact as @arceo2016 suggest on their work where invasion status and relatedness is considered. Notwithstanding, until our knowledge the effect of related and non-related species remains to be tested explicetely.

@@

The great environmental variability in natural systems and complexity of floral structures make heterospecific pollination studies a daunting task. For this reason, control pollination experiments where hand pollination is applied instead of using pollinators as vectors have simplified this task. However, other issues may arise such as

>>>>

Recent studies have advanced in the ecological understanding of heterospecific pollen effect @morales2008 @ashman2013 @arceo2016. A general overview of foreign pollen arrival is that it can play an important role on species fitness but seems to be context dependent and not always produce a decrease in fitness @morales2008. Part of this unpredictability is due to the enormous variability of foreing pollen transferred in nature, where levels between 0 and 75 percent are seen, but most commonly values ranges between 0 and 20 percent of the total pollen load, being the generalist species the ones that receive greater loads of heterospecific pollen @bartomeus2008 @montgomery2012 @ashman2013 @fang2013. Although heterospecific pollen quantity is fundamental to understand the outcome of the interaction so is the different traits of both pollen donor and recipient. @ashman2013 postulated the first predictive framework for traits of heterospecific pollen effect, where different traits such as compatibility system and pollen size among others seems to be crucial to understand foreing pollen effect. Moreover, in @tong2016 an assymetric effect was shown in a crossing experiment between 6 species of the genus *Pedicularis* where the pollen of long styled species was able to grow the full length of the style on short styled species but not viceversa. Despite these recent caveats, we still lack empirical evidence to affirm what are the main traits that drive heterospecific pollen effect for both pollen donor and recipient at seed production level. Interestingly, to comprehend how these traits interact is also crucial to look at the phylogenetic relatedness of the species. There is a considerable amount of literature of crosses between close related species @brown2001 @arceo2016can @tong2016 but few works focused on heterospecific pollen of far related species which is a more realistic feature in natural systems due to a less likely niche overlap (REF). Although close related species seems to affect greatly species fitness @arceo2016 the effect of heterospecific pollen of far and close related species remains to be tested explicitly.

@@

The great environmental variability in natural systems and complexity of floral structures make heterospecific pollination studies a daunting task. For this reason, controlled pollination experiments where hand pollination is applied instead of pollinators as pollen vectors have simplified this task. However, other issues may arise such as overestimation of the effect (REFs), standarized methods across studies or lack of descriptive methodology for reproducible work (REFs). One of the most common ways to estimate heterospecific pollen effect through hand pollination treatments is by using 50% pollen mixes of conspecific and heterospecific pollen which a realistic proportion (REF) but challenging because of the different pollen properties such as size or hydratation levels (REFS).

### 2018-12-04 Manuscript_draft/Manuscript_draft.tex
last edits of today

Recent studies have advanced in the ecological understanding of
heterospecific pollen effect Morales and Traveset (2008) Ashman and
Arceo-Gómez (2013) Arceo-Gómez and Ashman (2016). A general overview of
foreign pollen arrival is that it can play an important role on species
fitness but seems to be context dependent and not always produce a
decrease in fitness Morales and Traveset (2008). Part of this
unpredictability is due to the enormous variability of foreing pollen
transferred in nature, where levels between 0 and 75 percent are seen,
but most commonly values ranges between 10 and 20 percent of the total
pollen load, being the generalist species the ones that receive greater
loads of heterospecific pollen Montgomery and Rathcke (2012) Fang and
Huang (2013). Although heterospecific pollen quantity is fundamental to
understand the outcome of the interaction so is the different traits of
both pollen donor and recipient. Ashman and Arceo-Gómez (2013)
postulated the first predictive framework for traits of heterospecific
pollen effect, where different traits such as compatibility system and
pollen size among others seems to be crucial to understand foreing
pollen effect. Moreover, in Tong and Huang (2016) a assymetric effect
was shown in a crossing experiment between 6 species of the genus
\emph{Pedicularis} where the pollen of long styled species was able to
grow the full length of the style on short styled species but not
viceversa. Despite these recent caveats, we still lack empirical
evidence to affirm what are the main traits that drive heterospecific
pollen effect for both pollen donor and recipient at seed production
level. Interestingly, to comprehend how this traits interact is also
crucial to look at the phylogenetic relatedness of the species. There is
a considerable amount of literature of crosses between close related
species Brown and Mitchell (2001) Arceo-Gómez et al. (2016) Tong and
Huang (2016) for which is predicted to have a greater negative impact as
Arceo-Gómez and Ashman (2016) suggest on their work where invasion
status and relatedness is considered. Notwithstanding, until our
knowledge the effect of related and non-related species remains to be
tested explicetely.

@@

The great environmental variability in natural systems make
heterospecific pollination studies a daunting task. Moreover, when
pollinators are not used as pollen vectors, to control for standarized
pollen quantities hand pollination treatments are applied. However,
different methodologies across studies and different proxys of effect
difficulty comparisons across studies.

>>>>

Recent studies have advanced in the ecological understanding of
heterospecific pollen effect Morales and Traveset (2008) Ashman and
Arceo-Gómez (2013) Arceo-Gómez and Ashman (2016). A general overview of
foreign pollen arrival is that it can play an important role on species
fitness but seems to be context dependent and not always produce a
decrease in fitness Morales and Traveset (2008). Part of this
unpredictability is due to the enormous variability of foreing pollen
transferred in nature, where levels between 0 and 75 percent are seen,
but most commonly values ranges between 0 and 20 percent of the total
pollen load, being the generalist species the ones that receive greater
loads of heterospecific pollen Bartomeus et al. (2008) Montgomery and
Rathcke (2012) Ashman and Arceo-Gómez (2013) Fang and Huang (2013).
Although heterospecific pollen quantity is fundamental to understand the
outcome of the interaction so is the different traits of both pollen
donor and recipient. Ashman and Arceo-Gómez (2013) postulated the first
predictive framework for traits of heterospecific pollen effect, where
different traits such as compatibility system and pollen size among
others seems to be crucial to understand foreing pollen effect.
Moreover, in Tong and Huang (2016) an assymetric effect was shown in a
crossing experiment between 6 species of the genus \emph{Pedicularis}
where the pollen of long styled species was able to grow the full length
of the style on short styled species but not viceversa. Despite these
recent caveats, we still lack empirical evidence to affirm what are the
main traits that drive heterospecific pollen effect for both pollen
donor and recipient at seed production level. Interestingly, to
comprehend how these traits interact is also crucial to look at the
phylogenetic relatedness of the species. There is a considerable amount
of literature of crosses between close related species Brown and
Mitchell (2001) Arceo-Gómez et al. (2016) Tong and Huang (2016) but few
works focused on heterospecific pollen of far related species which is a
more realistic feature in natural systems due to a less likely niche
overlap (REF). Although close related species seems to affect greatly
species fitness Arceo-Gómez and Ashman (2016) the effect of
heterospecific pollen of far and close related species remains to be
tested explicitly.

@@

The great environmental variability in natural systems and complexity of
floral structures make heterospecific pollination studies a daunting
task. For this reason, controlled pollination experiments where hand
pollination is applied instead of pollinators as pollen vectors have
simplified this task. However, other issues may arise such as
overestimation of the effect (REFs), standarized methods across studies
or lack of descriptive methodology for reproducible work (REFs). One of
the most common ways to estimate heterospecific pollen effect through
hand pollination treatments is by using 50\% pollen mixes of conspecific
and heterospecific pollen which a realistic proportion (REF) but
challenging because of the different pollen properties such as size or
hydratation levels (REFS).
